Is machine learning good or bad for the natural sciences?

### 2023-12-31 paper/main.tex
references, tweaks

% To do:
% ------
% - Change it to be about PHYSICS rather than natural science?
% - Get feedback from Yann, BS, MSV, JH, others
% - put in all references
% - audit bibliography; make sure arXiv references in the reference list are consistent?
% - Publish on the arXiv.
% - set formatting and anonymization switches to go between arXiv and ICML Position Paper
% - submit to ICML.
% - Wait for the call from Stockholm.

@@

% Hogg typesetting issues
\setlength{\textwidth}{5.50in}
\setlength{\oddsidemargin}{0.00in}
\setlength{\oddsidemargin}{0.00in}
\setlength{\textheight}{9.25in}
\setlength{\topmargin}{-0.50in}
\setlength{\parindent}{1.2\baselineskip} % seriously
\renewcommand{\title}[1]{\section*{\raggedright #1}}
\renewcommand{\author}[1]{\medskip\par\noindent\textbf{#1}}
\newcommand{\affil}[1]{{\footnotesize\par\noindent #1 \par}}
\renewenvironment{abstract}{\paragraph{Abstract:}}{}
\newcommand{\documentname}{\textsl{Position Paper}}
\newcommand{\draftstatus}{\textcolor{red}{Draft version 2023-12-30}}
\newcommand{\sectionname}{Section}
\newenvironment{hoggnumerate}
  {\begin{list}
    {\arabic{enumii}.}
    {\usecounter{enumii}
     %\setlength{\labelwidth}{3em}
     %\setlength{\labelsep}{0em}
     \setlength{\itemsep}{0in}
     \setlength{\parsep}{0in}
     \setlength{\parskip}{0in}
     %\setlength{\leftmargin}{1.5cm}
     %\setlength{\rightmargin}{2cm}
     %\setlength{\itemindent}{0em} 
     %\let\makelabel=\makeboxlabel
    }
  }
{\end{list}}
\pagestyle{myheadings}
\markboth{foo}{\textcolor{gray}{\textsf{Hogg / Is machine learning good or bad for the natural sciences? \draftstatus}}}
\linespread{1.08}
\frenchspacing\raggedbottom\sloppy\sloppypar

@@

\paragraph{Our contributions:}
\begin{itemize}
  \item We deliver a description of the fundamental \emph{ontology} and \emph{epistemology} of machine learning, and contrast these with the ontologies and epistemologies of the natural sciences.
  \item We elucidate two important and strong statistical biases that are being introduced to the natural sciences by some of the uses of ML in those disciplines. One is a confirmation bias that arises when simulations are replaced or augmented by emulators. Another is a more standard estimator bias that is (possibly enormously) amplified when elements of datasets produced by ML regressions are used in combination or joint analyses.
  \item We show that there are many safe places for the use of ML methods in current natural-science practices, and places where (in the contemporary context) the use of ML methods is effectively \emph{required}. Many of these places are in the operational parts of scientific projects.
  \item Further, we argue that there are contexts inside a natural-scientific analysis in which using the most expressive ML method delivers \emph{the most conservative approach} to the scientific problem. An example is in causal contexts, where making a confounder model more expressive can strengthen conclusions about the causal channels of interest.
\end{itemize}

@@

  Below (\sectionname~\ref{sec:bad}) we argue that this is not an advisable use of ML in the natural sciences if the expectation is that the regression-generated labels are going to be used in downstream scientific projects.

@@

@@

@@

  HOGG: THERE ARE MANY different approaches to emulation; summarize here?
  Below (\sectionname~\ref{sec:bad}) we argue that the introduction of ML-based emulators can introduce unwanted confirmation bias.

@@

  When the goal is for these nuisances to be effectively modeled but not necessarily understood in detail, ML approaches can be effective and (as we argue in \sectionname~\ref{sec:good}) even conservative.

@@

@@

It doesn't mention using ML to emulate likelihoods or likelihood ratios \cite{biwei, kyle}.

@@

In Bayesian language, this problem is closely related to the point that when data are to be combined, they should be combined at the likelihood level, not the posterior level.
Likelihoods are not affected by population-level biases.
Likelihoods can produce unbiased point estimates.
Likelihood-based estimators are not usually the lowest-risk estimators, but they can be low- or zero-bias estimates.
The outputs of regressions are more like posterior-based estimators, affected by the implicit prior set by the training set.
When they are combined, the implicit prior gets amplified.
Most ML regressions are not capable of generating or emulating likelihoods.
There are interesting approaches in development to replace regressions with generative models that can produce likelihoods HOGG CITE THINGS.

@@

\paragraph{Emulator-induced confirmation bias:}
In many fields in the natural sciences, the fundamental theoretical model is \emph{computational}, meaning that theoretical predictions are made with large computer simulations.
These simulations are usually limited in the range of length scales and time scales which they can simultaneously compute.
Comparison of data with simulations usually involves running enormous numbers of simulations \cite{sbi}, because simulations have to span ranges of fundamental parameters and initial conditions and so on.
Thus these simulation requirements are getting very large for contemporary research projects.
For example, in cosmology contemporary experiments could easily consume the total computing capacity of the United States.
Hence we turn to emulators---trained ML regressions that have learned the input-output relationship of a simulation, or the relationship between an inexpensive low-resolution simulation and an expensive high-resolution experiment.
Emulators can make computationally impossible projects possible.

Recall that because these emulators are ML methods, they generally have uninterpretable internal parameters and weights, and they generally have been validated by comparison with held-out training data.
Like all high-capacity ML methods, they generally are vulnerable to adversarial attacks.

Now imagine that two cosmology experiments proceed to do enormous emulator-based inferences on large data sets.
Imagine that these inferences are so large that it would have been exceedingly expensive to have done them with the original first-principles simulations instead of the emulators.
Experiment A finds cosmological parameters very much in line with our expectations going in.
Experiment B finds something extremely surprising about the mass matrix for the neutrinos, in conflict with other measurements and our expectations.
The comment is made, when the results of Experiment B come out, that perhaps the anomaly comes from something very slightly wrong or off in the trained ML emulator.

@@

\section{When is machine learning good for natural science?}\label{sec:good}
In \sectionname~\ref{sec:intro} we re-phrased the question in the title as
``Where, in the natural sciences, can you use a model you don't understand?''
The answer is: You can use it for the parts of your scientific project or experiment or analysis that \emph{you don't need to understand}.
We have colleagues who believe that there is no such part.
We disagree: For example in astronomy, we use charge-coupled devices to make extremely precise measurements of stars, without knowing in detail the solid-state physics that makes them efficient, linear detectors of photons.
There are many parts of a scientific project that must work well, but which do not need to be carefully controlled or understood (in advance, anyway).

@@

\paragraph{Discoveries of outliers and rare objects:}
As mentioned in \sectionname~\ref{sec:why}, ML is well suited to outlier detection.
Since ML methods are expressive, and trained to fit the data, new data that \emph{aren't} well described by a trained ML model are possibly interesting.
Outlier-detection systems can be used to identify time intervals in which the equipment is malfunctioning.
They can also be used to identify objects that are rare or unusual.
Many discoveries in astrophysics are discoveries of new kinds of objects, originally found as data outliers \cite{quasars, voorwerp}.
So far, ML has not been involved in any big discoveries in astronomy, but we expect this to be a productive avenue for ML in the natural sciences and astronomy in particular.

\paragraph{Foreground, background, and confounder models:}
We mentioned in \sectionname~\ref{sec:philosophy} that a natural-science theory is a combination of a fundamental, latent theory with a (usually complicated) auxiliary observation model that explains the details of the observations.
The simplest part of this observation model is the model of the backgrounds, foregrounds, and confounders.
For example, any maps of the cosmic microwave background (by the ESA \textsl{Planck} spacecraft \cite{planck_maps}, for example) will have emission from the intervening Milky Way and from radio galaxies, adding to, and distorting the maps.
When the goal is to understand the cosmic microwave background, the investigator often doesn't need to understand these foregrounds \emph{physically}; an \emph{effective model} is sufficient.
And indeed, all the bleeding-edge methods for this problem currently are ML methods \cite{cmb_foregrounds}.
It doesn't matter to the questions of physical cosmology what those foregrounds \emph{mean}; and indeed with ML-based foreground removal, cosmic microwave background experiments have delivered the most precise measurements of the parameters of the cosmological model with sub-percent accuracy \cite{planck_parameters}.

This is all true: For foregrounds and backgrounds, understanding is not always necessary.
An even stronger statement can be made here, however, in the realm of causal inference:
Sometimes the goal is to argue that an effect is being caused by some particular cause, and there are confounding causal inputs that might be distorting or influencing the data.
The more expressive the model used to model the confounders, the stronger the conclusions that can be made about the causation \cite{causal_inference}.
That is, the huge capacity of ML methods can made a causal claim \emph{very conservative}, and hence very robust.
One way to see this is that if the confounder model has been given every chance it possibly can have to model the effect of interest, and it cannot, then it is a stronger conclusion that the effect is caused by the cause of interest.
This is the strategy in contemporary causal-inference projects, which make use of ML methods for these reasons \cite{bart}.

@@

That said, there is a certain amount of hype in this area, and we are seeing scientists ``move fast and break things'' \cite{zuck} all over the place.
If anything, this \documentname{} represents a call to slow down and think more.
ML tools are incredibly powerful, if we put them to use in the right places in the natural sciences.

There is a famous ancient quotation about ``the unreasonable effectiveness of mathematics in physics'' \cite{weiner_maybe}.
Mathematics has been pursued (in many cases) for abstract reasons of beauty and technical challenge, but it has ended up creating a language for many of the natural sciences.
Similarly, we could coin a phrase about \emph{the unreasonable effectiveness of machine learning in the natural sciences}.
It's unreasonable because ML methods were developed, engineered, and optimized to solve problems for commerce.
The most remarkable ML systems have been driven by critical industrial questions such as:
Which advertisement should I display on this web search result?
What direction should I steer this automobile?
How could I make a chair that looks like an avocado?
And which online videos contain kittens?
In some sense it is remarkable that there is \emph{any} overlap between these technologies and the things we need in scientific domains.

@@

To illustrate these ideas, we construct a toy data analysis.
This toy will make a general point, but it has a decidedly astronomical feel to it.
The complete toy data set is generated by a process with the algorithm below.
This algorithm makes references to objects $\xi_\eta$ and $\xi_\zeta$ which are generated prior to the start.
\begin{hoggnumerate}
    \item A floating-point value of a known parameter called ``guiding radius'' $r$ is generated in the range $0<r<14$.
    \item A floating-point value of a latent parameter called ``age'' $\eta$ is generated from a Gaussian with mean $14 - r$ and variance 4.
    \item The point is discarded if the age is outside the range $0<\eta<14$. If the point is not discarded:
    \item A $K$-dimensional latent vector $\zeta$ with $K=14$ is generated from a Gaussian with zero mean and variance 1.
    \item A $M$-dimensional latent vector $\xi$ with $M=110$ is created by $\xi = \xi_\eta\,\eta + \xi_\zeta\cdot\zeta$, where $\xi_\eta$ is a $M$-vector and $\xi_\zeta$ is a $M\times K$ matrix.
    The elements of $\xi_\eta$ and $\xi_\zeta$ were drawn (before the start) from Gaussians with zero mean and variances $0.01/K$ and $1/K$ respectively.
    \item A rectified latent vector $\tilde{\xi}$ is created by taking $1 + \xi$ and rectifying all pixels to lie in the range $[0, 1]$.
    \item A $M$-dimensional feature vector $x$ called ``the data'' is generated from a Gaussian with mean $\tilde{\xi}$ and variance $0.0025$.
    \item A label $y$ called ``measured age'' is generated from a Gaussian with mean $\eta$ and variance 1.
\end{hoggnumerate}
The data generated this way are shown in the top panels of \figurename~\ref{fig:regression}.
The data contain a non-trivial relationship between label $y$ (measured age) $y$ and known parameter $r$ (guiding radius).
This relationship is shown in \figurename~\ref{fig:regression}.
It isn't precisely linear because of the removal of data points with latent ages $\eta$ outside the range $0<\eta<14$.

We construct a three-layer multilayer perceptron (MLP) (using the scikit-learn implementation) with layer sizes of 64, 32, and 16 neurons.
The model is trained on a training set of 4096 data points with features $x$ and labels $y$.
The trained model takes as input features $x$ and outputs estimated labels $\hat{y}$.
The model is validated on a validation set of 2048 data points with features $x$ and labels $y$.
The test set is a much larger set of $10^5$ data points with features $x$ and no labels.
The model is used to make estimated labels $\hat{y}$ in the full test set.
The validation and test labels are shown in the bottom panels of \figurename~\ref{fig:regression}.
Importantly, \emph{the training, validation, and test samples are drawn from exactly the same process}, with the same distribution in all properties.
\begin{figure}[p!]
\includegraphics[width=0.49\textwidth]{notebooks/data_examples.png}
\includegraphics[width=0.49\textwidth]{notebooks/training_data.png} \\
\includegraphics[width=0.49\textwidth]{notebooks/validation.png}
\includegraphics[width=0.49\textwidth]{notebooks/test_data_results.png}
\caption{\sffamily%
Visualization of the toy regression. \textsl{Top-left:} Random examples of the data vectors $x$, which are one-dimensional images generated from a linear model plus a nonlinearity created by two rectifications. Details of the data generation are given in the text. Each example $x$ is labeled on the right side by the value of its label $y$. \textsl{Top-right:} The training-set labels $y$, plotted against the known parameter $r$, which is not used in the regression (only the vectors $x$ are used). Also shown are solid red circles showing the true mean relationship between $y$ and $r$ in the toy data. Open black squares show the empirical mean relationship measured in bins in the training-set data. \textsl{Bottom-left:} Validation of the trained regression in the validation set, showing that the label estimates $\hat{y}$ are noisy (as expected given the problem set-up) but not strongly biased. \textsl{Bottom-right:} The regression estimates $\hat{y}$ in the very large test set, plotted against the known parameter $r$. Also shown are the same solid red circles and open black squares as in the top-right plot. Black X-shaped symbols show the mean relationship between $\hat{y}$ and $r$. The relationship shown by the Xs is very precise but biased far away from the true relationship, unlike the relationship shown by the open squares (measured in the training set alone).\label{fig:regression}}
\end{figure}

The regression-estimated labels $\hat{y}$ are related to the measured labels $y$ in the held-out validation set with a linear relationship with unit slope and zero intercept.
The comparison to the validation data (bottom-left panel of \figurename~\ref{fig:regression}) shows that they are not precisely estimated, but there is no evidence for strong bias at the individual-object level.

The true relationship between label $y$ (measured age) and known parameter $r$ (guiding radius) is estimated in a large generated data set by taking means of $y$ in bins of $r$.
This relationship is shown in \figurename~\ref{fig:regression} with solid red circles.
The empirical relationship between label $y$ and known parameter $r$ in the training-set data is shown in \figurename~\ref{fig:regression} with open black squares.
These two relationships are statistically consistent, as expected.
The empirical relationship in the test-set data between the estimated label $\hat{y}$ and the known parameter $r$ in the test-set data is shown in \figurename~\ref{fig:regression} with black X-shaped symbols.
The test-set relationship is strongly biased, deviating from the true relationship at the edges of the $r$ range.

@@

In this case, what should the investigator do---the investigator who wants to know the relationship between $y$ (age) and $r$ (guiding radius)?
The investigator should \emph{just use the training set}.
Transferring labels to the test set did not help, even though the test set is far larger than the training set.
The test-set results are very precise (error bars on the X-shaped symbols cannot be visibly shown in \figurename~\ref{fig:regression}), but they are very wrong.
The training-set results are less precise but unbiased; they are much closer to the truth.

@@

Related to that: The most interesting aspect of this problem is that there is no \emph{wrong} bias:
The bias introduced from the training set is a correct bias; the training set represents a prior that accords with our beliefs about the test set.
And yet the population-level inference is biased.
The averaging of points shown by the X-shaped symbols in \figurename~\ref{fig:regression} is a simple misuse of the regression outputs; that point is trivial.
However, it is a misuse that is occurring repeatedly in astrophysics with catalogs generated by ML regressions HOGG CITE.

>>>>

% To do:
% ------
% - Change it to be about PHYSICS rather than natural science?
% - Get feedback from BS, MSV, JH, others
% - Update appendix with feedback from Vedant Chandra (email).
% - put in all references
% - audit bibliography; make sure arXiv references in the reference list are consistent? Also capitalization!
% - Publish on the arXiv.
% - set formatting and anonymization switches to go between arXiv and ICML Position Paper
% - submit to ICML.
% - Wait for the call from Stockholm.

@@

% Hogg typesetting issues that won't conflict with ICML
\frenchspacing
\newcommand{\documentname}{\textsl{Position Paper}}
\newcommand{\draftstatus}{\textcolor{red}{Draft version 2023-12-31}}
\newcommand{\sectionname}{Section}
\newcommand{\secref}[1]{\sectionname~\ref{#1}}
\newcommand{\figref}[1]{\figurename~\ref{#1}}
\newenvironment{hoggnumerate}
  {\begin{list}
    {\arabic{enumii}.}
    {\usecounter{enumii}
     %\setlength{\labelwidth}{3em}
     %\setlength{\labelsep}{0em}
     \setlength{\itemsep}{0in}
     \setlength{\parsep}{0in}
     \setlength{\parskip}{0in}
     %\setlength{\leftmargin}{1.5cm}
     %\setlength{\rightmargin}{2cm}
     %\setlength{\itemindent}{0em} 
     %\let\makelabel=\makeboxlabel
    }
  }
{\end{list}}

% Hogg typesetting issues that will conflict with ICML
\setlength{\textwidth}{5.50in}
\setlength{\oddsidemargin}{-0.25in}
\setlength{\textheight}{9.25in}
\setlength{\topmargin}{-0.50in}
\setlength{\parindent}{1.2\baselineskip} % seriously
\renewcommand{\title}[1]{\section*{\raggedright #1}}
\renewcommand{\author}[1]{\medskip\par\noindent\textbf{#1}}
\newcommand{\affil}[1]{{\footnotesize\par\noindent #1 \par}}
\renewenvironment{abstract}{\paragraph{Abstract:}}{}
\pagestyle{myheadings}
\markboth{foo}{\textcolor{gray}{\textsf{Hogg / Is machine learning good or bad for the natural sciences? \draftstatus}}}
\linespread{1.08}
\raggedbottom\sloppy\sloppypar

@@

\paragraph{Our contributions:}
\begin{itemize}
  \item We deliver a description of the fundamental \emph{ontology} and \emph{epistemology} of machine learning, and contrast these with the ontologies and epistemologies of the natural sciences.
  \item We elucidate two important and strong statistical biases that are being introduced to the natural sciences by some of the uses of ML in those disciplines. One is a confirmation bias that arises when simulations are replaced or augmented by emulators. Another is a more standard estimator bias that is (possibly enormously) amplified when elements of datasets produced by ML regressions are used in combination or joint analyses. These biases can't be easily corrected.
  \item We show that there are many safe places for the use of ML methods in current natural-science practices, and places where (in the contemporary context) the use of ML methods is effectively \emph{required}. Many of these places are in the operational parts of scientific projects.
  \item We argue that in causal contexts, in which the ML method is used to model foregrounds, backgrounds, calibration parameters, or other confounders, using the most expressive ML method can deliver \emph{the most conservative approach} to the scientific problem.
\end{itemize}

@@

  Below (\secref{sec:bad}) we argue that this is not an advisable use of ML in the natural sciences if the expectation is that the regression-generated labels are going to be used in downstream scientific projects.

@@

  It is subject to the same kinds of biases described below (\secref{sec:bad}).

@@

  It is critical, if ML methods are used for real-time operations, that the methods and all their latent parameters (weights) be preserved for analysis, operations simulations, and conterfactual exercises, often performed long after the data are taken.

@@

  Below (\secref{sec:bad}) we argue that the introduction of ML emulators can introduce an unwanted, strong confirmation bias.

@@

  When the goal is for these nuisances to be effectively modeled but not necessarily understood in detail, ML approaches can be effective and (as we argue in \secref{sec:good}) even conservative.

@@

  So far this hope has borne no important new discoveries, but discoveries are not inconceivable.

@@

It doesn't mention using ML to emulate likelihoods or likelihood ratios \cite{biwei, likelihood_ratio}.

@@

In Bayesian language, this problem is closely related to the point that when data are to be combined, they should be combined at the likelihood level, not the posterior level.
Likelihoods are not affected by population-level biases.
Likelihoods can produce unbiased point estimates.
Likelihood-based estimators are not usually the lowest-risk estimators, but they can be low- or zero-bias estimates.
The outputs of regressions are more like posterior-based estimators, affected by the implicit prior set by the training set.
When they are combined, the implicit prior gets amplified.
Most ML regressions are not capable of generating or emulating likelihoods.
There are interesting approaches in development to replace standard regressions with generative models that can produce likelihoods \cite{cannon, biwei, likelihood_ratio}.

@@

\paragraph{Emulator-induced confirmation bias:}
In many fields in the natural sciences, the fundamental theoretical model is \emph{computational}, meaning that theoretical predictions are made with large computer simulations.
These simulations are usually being asked to handle large ranges of length scales and time scales, so they are generally getting larger and more computationally demanding each year.
Quantitative comparison of data with simulations usually involves running enormous numbers of simulations \cite{abc}, because simulations have to span ranges of fundamental parameters and initial conditions and so on.
Thus these simulation requirements are getting very large for contemporary research projects.
For example, in cosmology, contemporary experiments could easily consume the total computing capacity of the United States.
Hence we turn to emulators---trained ML regressions that have learned the input-output relationship of a simulation, or the relationship between an inexpensive low-resolution simulation and an expensive high-resolution experiment.
Emulators can make computationally impossible projects possible.

Recall that because these emulators are ML methods, they generally have uninterpretable internal parameters and weights, and they generally have been validated by comparison with held-out training data.
Like all high-capacity ML methods, they generally are vulnerable to adversarial attacks \cite{adversarial1}.

Now imagine that two cosmology experiments proceed to do enormous emulator-based inferences on large data sets.
Imagine that these inferences are so large that it would have been exceedingly expensive to have done them with the original first-principles simulations instead of the emulators.
Experiment A finds cosmological parameters very much in line with our expectations going in.
Experiment B finds something extremely surprising about the mass matrix for the neutrinos, in conflict with other measurements and our expectations.
Is it possible that the anomaly discovered by Experiment B isn't real; it just comes from something very slightly wrong or off in the trained ML emulator?

@@

\section{When is machine learning good for natural science?}\label{sec:good}
In \secref{sec:intro} we re-phrased the question in the title as
``Where, in the natural sciences, can you use a model you don't understand?''
The answer is: You can use it for the parts of your scientific project or experiment or analysis that \emph{you don't need to understand}.
We have colleagues who believe that there is no such part.
We disagree: For example in astronomy, we use charge-coupled devices to make extremely precise measurements of stars, without knowing in detail the solid-state physics that makes them efficient, linear detectors of photons.
There are many parts of a scientific project that must work well, but which do not need to be carefully controlled or understood (in advance, anyway).

@@

\paragraph{Discoveries of outliers and rare objects:}
As mentioned in \secref{sec:why}, ML is well suited to outlier detection.
Since ML methods are expressive, and trained to fit the data, new data that \emph{aren't} well described by a trained ML model are possibly interesting.
Outlier-detection systems can be used to identify time intervals in which the equipment is malfunctioning.
They can also be used to identify objects that are rare or unusual.
Many discoveries in astrophysics are discoveries of new kinds of objects, originally found as data outliers \cite{quasars, voorwerp}.
So far, ML has not been involved in any big discoveries in astronomy, but we expect this to be a productive avenue for ML in the natural sciences and astronomy in particular.

\paragraph{Foreground, background, and confounder models:}
We mentioned in \secref{sec:philosophy} that a natural-science theory is a combination of a fundamental, latent theory with a (usually complicated) auxiliary observation model that explains the details of the observations.
The simplest part of this observation model is the model of the backgrounds, foregrounds, and confounders.
For example, any maps of the cosmic microwave background (by the ESA \textsl{Planck} spacecraft \cite{planck_maps}, for example) will have emission from the intervening Milky Way and from radio galaxies, adding to, and distorting the maps.
When the goal is to understand the cosmic microwave background, the investigator often doesn't need to understand these foregrounds \emph{physically}; an \emph{effective model} is sufficient.
And indeed, all the bleeding-edge methods for this problem currently are ML methods \cite{cmb_foregrounds}.
It doesn't matter to the questions of physical cosmology what those foregrounds \emph{mean}; and indeed with ML-based foreground removal, cosmic microwave background experiments have delivered the most precise measurements of the parameters of the cosmological model with sub-percent accuracy \cite{planck_parameters}.
For foregrounds and backgrounds, understanding is not always necessary.

An even stronger statement can be made here, however, in the realm of causal inference:
Sometimes the goal is to argue that an effect is being caused by some particular cause, and there are confounding causal inputs that might be distorting or influencing the data.
The more expressive the model used to model the confounders, the stronger the conclusions that can be made about the causation \cite{causal_inference}.
That is, the huge capacity of ML methods can made a causal claim \emph{very conservative}, and hence very robust.
One way to see this is that if the confounder model has been given every chance it possibly can have to model the effect of interest, and it cannot, then it is a stronger conclusion that the effect is caused by the cause of interest.
This is the strategy in contemporary causal-inference projects, which make use of ML methods for these reasons \cite{bart}.

@@

That said, there is a certain amount of hype in this area, and we are seeing scientists ``move fast and break things'' all over the place.
If anything, this \documentname{} represents a call to slow down and think more.
ML tools are incredibly powerful, if we put them to use in the right places in the natural sciences.
In particular, the biases we highlight in \secref{sec:bad}

There is a famous article by Wigner about ``the unreasonable effectiveness of mathematics in the natural sciences'' \cite{wigner}.
Mathematics has been pursued (in many cases) for abstract reasons of beauty and technical challenge, but it has ended up creating a language for many of the natural sciences.
Similarly, we could coin a phrase about \emph{the unreasonable effectiveness of machine learning in the natural sciences}.
It's unreasonable because ML methods were developed, engineered, and optimized to solve problems for commerce.
The most remarkable ML systems have been driven by critical industrial questions such as:
Which advertisement should I display on this web search result?
What direction should I steer this automobile?
How could I make a chair that looks like an avocado?
And which online videos contain kittens?
In some sense it is remarkable that there is \emph{any} overlap between these technologies and the things we need in scientific domains.

@@

To illustrate these ideas, we construct a toy data analysis.
This toy will make a general point, but it has a decidedly astronomical feel to it.
The complete toy data set is generated by a process with the algorithm below.
This algorithm makes references to objects $\xi_\eta$ and $\xi_\zeta$ which are generated prior to the start.
\begin{hoggnumerate}
    \item A floating-point value of a known parameter called ``guiding radius'' $r$ is generated in the range $0<r<14$.
    \item A floating-point value of a latent parameter called ``age'' $\eta$ is generated from a Gaussian with mean $14 - r$ and variance 4.
    \item The point is discarded if the age is outside the range $0<\eta<14$. If the point is not discarded:
    \item A $K$-dimensional latent vector $\zeta$ with $K=14$ is generated from a Gaussian with zero mean and variance 1.
    \item A $M$-dimensional latent vector $\xi$ with $M=110$ is created by $\xi = \xi_\eta\,\eta + \xi_\zeta\cdot\zeta$, where $\xi_\eta$ is a $M$-vector and $\xi_\zeta$ is a $M\times K$ matrix.
    The elements of $\xi_\eta$ and $\xi_\zeta$ were drawn (before the start) from Gaussians with zero mean and variances $0.01/K$ and $1/K$ respectively.
    \item A rectified latent vector $\tilde{\xi}$ is created by taking $1 + \xi$ and rectifying all pixels to lie in the range $[0, 1]$.
    \item A $M$-dimensional feature vector $x$ called ``the data'' is generated from a Gaussian with mean $\tilde{\xi}$ and variance $0.0025$.
    \item A label $y$ called ``measured age'' is generated from a Gaussian with mean $\eta$ and variance 1.
\end{hoggnumerate}
The data generated this way are shown in the top panels of \figref{fig:regression}.
The data contain a non-trivial relationship between label $y$ (measured age) $y$ and known parameter $r$ (guiding radius).
This relationship is shown in \figref{fig:regression}.
It isn't precisely linear because of the removal of data points with latent ages $\eta$ outside the range $0<\eta<14$.

We construct a three-layer multilayer perceptron (MLP) (using the scikit-learn implementation) with layer sizes of 64, 32, and 16 neurons.
The model is trained on a training set of 4096 data points with features $x$ and labels $y$.
The trained model takes as input features $x$ and outputs estimated labels $\hat{y}$.
The model is validated on a validation set of 2048 data points with features $x$ and labels $y$.
The test set is a much larger set of $10^5$ data points with features $x$ and no labels.
The model is used to make estimated labels $\hat{y}$ in the full test set.
The validation and test labels are shown in the bottom panels of \figref{fig:regression}.
Importantly, \emph{the training, validation, and test samples are drawn from exactly the same process}, with the same distribution in all properties.
\begin{figure}[p!]
\includegraphics[width=0.49\textwidth]{notebooks/data_examples.png}
\includegraphics[width=0.49\textwidth]{notebooks/training_data.png} \\
\includegraphics[width=0.49\textwidth]{notebooks/validation.png}
\includegraphics[width=0.49\textwidth]{notebooks/test_data_results.png}
\caption{\sffamily%
Visualization of the toy regression. \textsl{Top-left:} Random examples of the data vectors $x$, which are one-dimensional images generated from a linear model plus a nonlinearity created by two rectifications. Details of the data generation are given in the text. Each example $x$ is labeled on the right side by the value of its label $y$. \textsl{Top-right:} The training-set labels $y$, plotted against the known parameter $r$, which is not used in the regression (only the vectors $x$ are used). Also shown are solid red circles showing the true mean relationship between $y$ and $r$ in the toy data. Open black squares show the empirical mean relationship measured in bins in the training-set data. \textsl{Bottom-left:} Validation of the trained regression in the validation set, showing that the label estimates $\hat{y}$ are noisy (as expected given the problem set-up) but not strongly biased. \textsl{Bottom-right:} The regression estimates $\hat{y}$ in the very large test set, plotted against the known parameter $r$. Also shown are the same solid red circles and open black squares as in the top-right plot. Black X-shaped symbols show the mean relationship between $\hat{y}$ and $r$. The relationship shown by the Xs is very precise but biased far away from the true relationship, unlike the relationship shown by the open squares (measured in the training set alone).\label{fig:regression}}
\end{figure}

The regression-estimated labels $\hat{y}$ are related to the measured labels $y$ in the held-out validation set with a linear relationship with unit slope and zero intercept.
The comparison to the validation data (bottom-left panel of \figref{fig:regression}) shows that they are not precisely estimated, but there is no evidence for strong bias at the individual-object level.

The true relationship between label $y$ (measured age) and known parameter $r$ (guiding radius) is estimated in a large generated data set by taking means of $y$ in bins of $r$.
This relationship is shown in \figref{fig:regression} with solid red circles.
The empirical relationship between label $y$ and known parameter $r$ in the training-set data is shown in \figref{fig:regression} with open black squares.
These two relationships are statistically consistent, as expected.
The empirical relationship in the test-set data between the estimated label $\hat{y}$ and the known parameter $r$ in the test-set data is shown in \figref{fig:regression} with black X-shaped symbols.
The test-set relationship is strongly biased, deviating from the true relationship at the edges of the $r$ range.

@@

In this case, what should the investigator do---the investigator who wants to know the relationship between $y$ (age) and $r$ (guiding radius)?
The investigator should \emph{just use the training set}.
Transferring labels to the test set did not help, even though the test set is far larger than the training set.
The test-set results are very precise (error bars on the X-shaped symbols cannot be visibly shown in \figref{fig:regression}), but they are very wrong.
The training-set results are less precise but unbiased; they are much closer to the truth.

@@

Related to that: The most interesting aspect of this problem is that there is no \emph{wrong} bias:
The bias introduced from the training set is a correct bias; the training set represents a prior that accords with our beliefs about the test set.
And yet the population-level inference is biased.
The averaging of points shown by the X-shaped symbols in \figref{fig:regression} is a simple misuse of the regression outputs; that point is trivial.
However, it is a misuse that is occurring repeatedly in astrophysics with catalogs generated by ML regressions HOGG CITE.
